---
title: "APR Data Cleaning: Region Assignment and Raw Variable Extraction"
author: "WCER School Mental Health Collaborative"
date: "`r Sys.Date()`"
output:
  html_notebook:
    df_print: paged
    highlight: pygments
    number_sections: yes
    theme: default
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
  html_document:
    df_print: paged
    highlight: pygments
    number_sections: yes
    theme: default
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

## Overview

This notebook cleans and prepares Annual Performance Report (APR) data for the two
federal school mental health grant programs administered by the WCER School Mental
Health Collaborative: **MHSP** (Mental Health Service Professionals) and **SBMH**
(School-Based Mental Health). The source file is a two-sheet Excel workbook exported
from the APR data system.

The cleaning steps are:

1. Import each program sheet and standardize headers
2. Identify and resolve duplicate column names
3. Check for duplicate grantee records
4. Assign HHS geographic regions to each grantee based on state
5. Extract the raw-number GPRA outcome variables needed for downstream analysis
6. Export cleaned files to Excel

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
# Install any missing packages, then load all required libraries
packages <- c("readxl", "tidyverse", "writexl")
install.packages(setdiff(packages, installed.packages()[, "Package"]),
                 quiet = TRUE)
invisible(sapply(packages, library, character.only = TRUE))
```

---

## 2. MHSP Program Data

### 2.1 Import

The APR workbook uses a two-row header structure: row 1 is a metadata banner and
row 2 contains the actual column names. Row 3 is a formatting separator that must
also be dropped before analysis.

```{r mhsp-import, message=FALSE}
mhsp <- read_excel(
  "I:/GitHub Documents/2024/METRICS/Ad Hoc Requests/APR_Application_Data_Synthetic.xlsx",
  sheet = 1,
  col_names = FALSE
)

# Promote row 2 to column names, then drop header rows 1-3
colnames(mhsp) <- as.character(unlist(mhsp[2, ]))
mhsp <- mhsp[-c(1:3), ]

# Drop any trailing all-NA column produced by the Excel export
mhsp <- mhsp[, colSums(is.na(mhsp)) < nrow(mhsp)]

cat("MHSP rows after import:", nrow(mhsp), "\n")
cat("MHSP columns after import:", ncol(mhsp), "\n")
```

```{r mhsp-preview}
head(mhsp)
```

### 2.2 Resolve Duplicate Column Names

The APR export sometimes produces identically named columns when a GPRA measure
appears in both an annual and a current-year block. This function detects duplicated
names, verifies whether the two versions carry identical values, renames them with
`_1` / `_2` suffixes for inspection, and logs the result.

```{r mhsp-dup-cols}
resolve_duplicate_columns <- function(df) {
  dup_names <- unique(names(df)[duplicated(names(df))])

  if (length(dup_names) == 0) {
    message("No duplicate column names found.")
    return(df)
  }

  for (nm in dup_names) {
    idx <- which(names(df) == nm)

    if (length(idx) != 2) {
      warning(sprintf(
        "Expected exactly 2 copies of '%s' but found %d. Skipping.",
        nm, length(idx)
      ))
      next
    }

    # Rename for side-by-side comparison
    names(df)[idx] <- paste0(nm, "_", 1:2)

    identical_vals <- all(df[[idx[1]]] == df[[idx[2]]], na.rm = TRUE)
    message(sprintf(
      "Duplicate '%s' (cols %d & %d): values are %s.",
      nm, idx[1], idx[2],
      ifelse(identical_vals, "IDENTICAL", "DIFFERENT -- review required")
    ))
  }

  df
}

mhsp <- resolve_duplicate_columns(mhsp)
```

**Cross-check duplicated columns before dropping.** If `resolve_duplicate_columns()`
found duplicates, it renamed them to `_1` / `_2`. The checks below only run when those
suffixed columns exist; they should return zero rows if the values are truly identical.

```{r mhsp-crosscheck}
crosscheck_pair <- function(df, col1, col2) {
  if (!all(c(col1, col2) %in% names(df))) {
    message(sprintf("Columns '%s' and '%s' not present -- no duplicate to check.", col1, col2))
    return(invisible(NULL))
  }
  result <- df |>
    select(Grantee_Number, all_of(c(col1, col2))) |>
    filter(.data[[col1]] != .data[[col2]])
  if (nrow(result) == 0) {
    message(sprintf("'%s' vs '%s': values are identical across all rows.", col1, col2))
  } else {
    message(sprintf("'%s' vs '%s': %d row(s) differ -- review before dropping.", col1, col2, nrow(result)))
    print(result)
  }
}

crosscheck_pair(mhsp, "GPRA2_A_Annual_Ratio_Actual_DEN_1",       "GPRA2_A_Annual_Ratio_Actual_DEN_2")
crosscheck_pair(mhsp, "GPRA2_A_Annual_Ratio_Actual_NUM_1",       "GPRA2_A_Annual_Ratio_Actual_NUM_2")
crosscheck_pair(mhsp, "GPRA3_B_PreviousAnnual_Ratio_Target_1",   "GPRA3_B_PreviousAnnual_Ratio_Target_2")
```

Keep the first copy of each duplicated column and assign clean names. The rename
and drop only execute when the suffixed columns are present.

```{r mhsp-rename-cols}
# Rename _1 copies to their clean names (only if duplicates were found)
rename_map <- c(
  GPRA2_A_Annual_Ratio_Actual_DEN = "GPRA2_A_Annual_Ratio_Actual_DEN_1",
  GPRA2_A_Annual_Ratio_Actual_NUM = "GPRA2_A_Annual_Ratio_Actual_NUM_1",
  GPRA2_B_Annual_Ratio_Actual_NUM = "GPRA2_A_Annual_Ratio_Actual_NUM_2"
)
present_renames <- rename_map[rename_map %in% names(mhsp)]
if (length(present_renames) > 0) {
  mhsp <- rename(mhsp, all_of(present_renames))
}

# Drop the _2 redundant DEN column if it still exists
if ("GPRA2_A_Annual_Ratio_Actual_DEN_2" %in% names(mhsp)) {
  mhsp <- select(mhsp, -GPRA2_A_Annual_Ratio_Actual_DEN_2)
}

message(sprintf("MHSP column count after deduplication: %d", ncol(mhsp)))
```

### 2.3 Check for Duplicate Grantee Records

Each grantee should appear exactly once. Any duplicates must be resolved before
the file is used in analysis.

```{r mhsp-dup-rows}
mhsp |>
  group_by(Grantee_Number) |>
  filter(n() > 1) |>
  arrange(Grantee_Number)
```

### 2.4 Assign HHS Geographic Regions

Grantees are assigned to one of four broad geographic regions based on their
state abbreviation. These regions are used to stratify reporting and visualizations
in downstream analyses.

```{r region-setup}
# State-to-region lookup vectors
west      <- c("WA", "OR", "NV", "UT", "ID", "MT", "WY", "CO",
                "AZ", "NM", "AK", "CA", "HI")
central   <- c("ND", "MN", "WI", "MI", "NE", "SD", "KS", "IA",
                "OK", "MO", "IL", "IN")
northeast <- c("KY", "WV", "VA", "OH", "PA", "DC", "MD", "DE",
                "NJ", "RI", "CT", "MA", "NY", "VT", "NH", "ME")
southeast <- c("TX", "LA", "AR", "TN", "MS", "AL", "GA",
                "NC", "SC", "FL", "PR")

assign_region <- function(state) {
  dplyr::case_when(
    state %in% west      ~ "West",
    state %in% central   ~ "Central",
    state %in% northeast ~ "Northeast",
    state %in% southeast ~ "Southeast",
    TRUE                 ~ "Unknown"
  )
}
```

```{r mhsp-region}
mhsp <- mhsp |>
  mutate(Region = assign_region(State))

# Verify all states mapped correctly
mhsp |>
  group_by(Region) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(pct = round(n / sum(n) * 100, 1))
```

---

## 3. SBMH Program Data

### 3.1 Import

```{r sbmh-import, message=FALSE}
sbmh <- read_excel(
  "I:/GitHub Documents/2024/METRICS/Ad Hoc Requests/APR_Application_Data_Synthetic.xlsx",
  sheet = 2,
  col_names = FALSE
)

colnames(sbmh) <- as.character(unlist(sbmh[2, ]))
sbmh <- sbmh[-c(1:3), ]
sbmh <- sbmh[, colSums(is.na(sbmh)) < nrow(sbmh)]

cat("SBMH rows after import:", nrow(sbmh), "\n")
cat("SBMH columns after import:", ncol(sbmh), "\n")
```

```{r sbmh-preview}
head(sbmh)
```

### 3.2 Check for Duplicate Column Names

```{r sbmh-dup-cols}
dup_names <- names(sbmh)[duplicated(names(sbmh))]

if (length(dup_names) == 0) {
  message("No duplicate column names found in SBMH.")
} else {
  are_identical <- all(sbmh[[dup_names[1]]] == sbmh[[dup_names[2]]], na.rm = TRUE)
  message(sprintf(
    "Duplicate '%s': values are %s.",
    dup_names[1],
    ifelse(are_identical, "identical", "DIFFERENT -- review required")
  ))
}
```

### 3.3 Check for Duplicate Grantee Records

```{r sbmh-dup-rows}
sbmh |>
  group_by(Grantee_Number) |>
  filter(n() > 1) |>
  arrange(Grantee_Number)
```

### 3.4 Assign HHS Geographic Regions

```{r sbmh-region}
sbmh <- sbmh |>
  mutate(Region = assign_region(State))

# Check for any unmatched states
unknown_states <- sbmh |> filter(Region == "Unknown")

if (nrow(unknown_states) > 0) {
  message(sprintf("%d rows have an unrecognized state code:", nrow(unknown_states)))
  print(unknown_states |> select(Grantee_Number, State, Region))
} else {
  message("All states mapped to a region successfully.")
}
```

```{r sbmh-drop-na-grantee}
# Drop any rows where Grantee_Number is NA (formatting artifacts from the export)
sbmh <- sbmh |> filter(!is.na(Grantee_Number))

# Final region distribution
sbmh |>
  group_by(Region) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(pct = round(n / sum(n) * 100, 1))
```

---

## 4. Export Cleaned Files

### 4.1 MHSP -- Full Cleaned File and Raw-Variable Subset

```{r mhsp-export}
mhsp_output <- "I:/GitHub Documents/2024/METRICS/Ad Hoc Requests/mhsp_cleaned.xlsx"
write_xlsx(mhsp, mhsp_output)
cat("Full MHSP data saved to:", mhsp_output, "\n")

# Raw outcome variables only
mhsp_raw_vars <- c(
  "Grantee_Number", "Grantee_Name", "Grantee_Type", "State", "Region",
  "SUD_reference", "CP1_YN", "CP1_Met",
  "GPRA1_A_Annual_Raw_Target", "GPRA1_A_Annual_Raw_Actual",
  "GPRA1_B_Current_Raw_Target", "GPRA1_B_Current_Raw_Actual",
  "GPRA2_A_Annual_Raw_Target", "GPRA2_A_Annual_Raw_Actual",
  "GPRA2_B_Current_Raw_Target", "GPRA2_B_Current_Raw_Actual",
  "GPRA3_A_Annual_Raw_Target", "GPRA3_A_Annual_Raw_Actual",
  "GPRA3_B_PreviousAnnual_Raw_Target", "GPRA3_B_PreviousAnnual_Raw_Actual",
  "GPRA4_ PP1_Raw_Target", "GPRA4_ PP1_Raw_Actual"
)

mhsp_raw <- mhsp[, intersect(mhsp_raw_vars, names(mhsp))]
mhsp_raw_output <- "I:/GitHub Documents/2024/METRICS/Ad Hoc Requests/mhsp_raw.xlsx"
write_xlsx(mhsp_raw, mhsp_raw_output)
cat("MHSP raw-variable subset saved to:", mhsp_raw_output, "\n")
```

### 4.2 SBMH -- Full Cleaned File and Raw-Variable Subset

```{r sbmh-export}
sbmh_output <- "I:/GitHub Documents/2024/METRICS/Ad Hoc Requests/sbmh_cleaned.xlsx"
write_xlsx(sbmh, sbmh_output)
cat("Full SBMH data saved to:", sbmh_output, "\n")

# Raw outcome variables only
sbmh_raw_vars <- c(
  "Grantee_Number", "Grantee_Name", "Grantee_Type", "State", "Region",
  "SUD_Reference", "CP1_YN", "CP1_Met",
  "GPRA1_Raw_Target", "GPRA1_Raw_Actual",
  "GPRA2_Raw_Target", "GPRA2_Raw_Actual",
  "GPRA3_Ratio_Target", "GPRA3_Ratio_Actual",
  "GPRA4_Raw_Target", "GPRA4_Raw_Actual",
  "GPRA5_Raw_Target", "GPRA5_Raw_Actual",
  "GPRA6_Raw_Target", "GPRA6_Raw_Actual"
)

sbmh_raw <- sbmh[, intersect(sbmh_raw_vars, names(sbmh))]
sbmh_raw_output <- "I:/GitHub Documents/2024/METRICS/Ad Hoc Requests/sbmh_raw.xlsx"
write_xlsx(sbmh_raw, sbmh_raw_output)
cat("SBMH raw-variable subset saved to:", sbmh_raw_output, "\n")
```

---

## 5. Session Info

```{r session-info}
sessionInfo()
```
