---
title: "IPR Aggregated Data -- Raw GPRA Targets and Actuals"
author: "WCER School Mental Health Collaborative"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook:
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  html_document:
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes
---

## Overview

This notebook aggregates Interim Performance Report (IPR) data for the two federal
school mental health grant programs -- **MHSP** and **SBMH** -- and computes
program-wide totals and descriptive statistics for all GPRA outcome measures.

The workflow for each program:

1. Import the revised IPR Excel file and standardize headers
2. Detect and remove duplicate Grant IDs
3. Cross-check grant roster against the APR reference file
4. Clean each GPRA column (rename, coerce to numeric, recode 999 to NA)
5. Compute sum, mean, min, max, and n for each measure
6. Flag grantees with missing values
7. Produce a narrative "Additional Information" paragraph for each measure
8. Export a summary table and a revision follow-up list

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
source("R/helpers.R")
load_packages(c("readxl", "tidyverse", "writexl", "knitr"))
```

---

## 2. MHSP

### 2.1 Import

```{r mhsp-import}
mhsp <- import_ipr_sheet("data/REVISED_MHSP_IPR_Synthetic.xlsx")
cat("MHSP rows:", nrow(mhsp), "| Columns:", ncol(mhsp), "\n")
head(mhsp)
```

### 2.2 Duplicate Grant IDs

```{r mhsp-dupes}
mhsp <- deduplicate_grants(mhsp, id_col = "Grant ID")
```

### 2.3 Cross-check Against APR Roster

```{r mhsp-crosscheck, message=FALSE}
mhsp_apr <- suppressMessages(
  read_excel("data/APRs_Revised_Tiers_Synthetic.xlsx", sheet = 1)
)
cat("APR reference rows (MHSP):", nrow(mhsp_apr), "\n")
crosscheck_grants(mhsp$`Grant ID`, mhsp_apr$grantee_number,
                  ipr_label = "MHSP IPR", apr_label = "MHSP APR")
```

---

## 3. MHSP GPRA Measures

For each measure the steps are: rename, coerce to numeric, recode 999 to NA,
then print the sum/mean and list any grantees with missing values.

```{r mhsp-clean-gpra, warning=FALSE}
# Rename all GPRA columns in one pass
mhsp <- mhsp |>
  clean_gpra_col("GPRA 1A Target (Raw)", "gpra_1a_target") |>
  clean_gpra_col("GPRA 1A Actual (Raw)", "gpra_1a_actual") |>
  clean_gpra_col("GPRA 1B Target (Raw)", "gpra_1b_target") |>
  clean_gpra_col("GPRA 1B Actual (Raw)", "gpra_1b_actual") |>
  clean_gpra_col("GPRA 2A Target (Raw)", "gpra_2a_target") |>
  clean_gpra_col("GPRA 2A Actual (Raw)", "gpra_2a_actual") |>
  clean_gpra_col("GPRA 2B Target (Raw)", "gpra_2b_target") |>
  clean_gpra_col("GPRA 2B Actual (Raw)", "gpra_2b_actual") |>
  clean_gpra_col("GPRA 3A Target (Raw)", "gpra_3a_target") |>
  clean_gpra_col("GPRA 3A Actual (Raw)", "gpra_3a_actual") |>
  clean_gpra_col("GPRA 3B Target (Raw)", "gpra_3b_target") |>
  clean_gpra_col("GPRA 3B Actual (Raw)", "gpra_3b_actual")
```

### 3.1 GPRA 1 -- Providers Trained

#### Target

```{r mhsp-gpra1-target}
summarise_gpra_col(mhsp, "gpra_1a_target", "GPRA 1A Target")
missing_grantees(mhsp, "gpra_1a_target")
summarise_gpra_col(mhsp, "gpra_1b_target", "GPRA 1B Target")
missing_grantees(mhsp, "gpra_1b_target")
```

#### Actual

```{r mhsp-gpra1-actual}
summarise_gpra_col(mhsp, "gpra_1a_actual", "GPRA 1A Actual")
missing_grantees(mhsp, "gpra_1a_actual")
summarise_gpra_col(mhsp, "gpra_1b_actual", "GPRA 1B Actual")
missing_grantees(mhsp, "gpra_1b_actual")
```

#### Narrative

```{r mhsp-gpra1-narrative}
total_grantees <- nrow(mhsp)
submitted_1a   <- sum(!is.na(mhsp$gpra_1a_actual))
enrolled_1b    <- sum(mhsp$gpra_1b_target, na.rm = TRUE)

rev_summary <- revision_summary(mhsp, label = "MHSP")
rev_pct     <- rev_summary$Pct_Submitted

cat(glue::glue(
  "Additional Information: The data included in this report was gathered via 2024 ",
  "Interim Performance Reports and reflect interim performance based on {submitted_1a} ",
  "out of {total_grantees} grantees that submitted data for this measure. Grantees who ",
  "did not submit data have been contacted and provided with feedback. {rev_pct}% of ",
  "grantees who required revisions submitted revisions prior to 12/9/24 and were included ",
  "in this report.\n\nThe actual data reflects the cumulative number of school-based mental ",
  "health service providers who completed training by the grantee during the reporting period. ",
  "There are {enrolled_1b} (GPRA 1B) providers currently enrolled in training.\n\n",
  "Set targets for future years are based on actual data collected in year one."
))
```

### 3.2 GPRA 2 -- Providers in Placement

#### Target

```{r mhsp-gpra2-target}
summarise_gpra_col(mhsp, "gpra_2a_target", "GPRA 2A Target")
missing_grantees(mhsp, "gpra_2a_target")
summarise_gpra_col(mhsp, "gpra_2b_target", "GPRA 2B Target")
missing_grantees(mhsp, "gpra_2b_target")
```

#### Actual

```{r mhsp-gpra2-actual}
summarise_gpra_col(mhsp, "gpra_2a_actual", "GPRA 2A Actual")
missing_grantees(mhsp, "gpra_2a_actual")
summarise_gpra_col(mhsp, "gpra_2b_actual", "GPRA 2B Actual")
missing_grantees(mhsp, "gpra_2b_actual")
```

### 3.3 GPRA 3 -- Providers Hired and Retained

#### Target

```{r mhsp-gpra3-target}
summarise_gpra_col(mhsp, "gpra_3a_target", "GPRA 3A Target")
missing_grantees(mhsp, "gpra_3a_target")
summarise_gpra_col(mhsp, "gpra_3b_target", "GPRA 3B Target")
missing_grantees(mhsp, "gpra_3b_target")
```

#### Actual

```{r mhsp-gpra3-actual}
summarise_gpra_col(mhsp, "gpra_3a_actual", "GPRA 3A Actual")
missing_grantees(mhsp, "gpra_3a_actual")
summarise_gpra_col(mhsp, "gpra_3b_actual", "GPRA 3B Actual")
missing_grantees(mhsp, "gpra_3b_actual")
```

### 3.4 MHSP Summary Table

```{r mhsp-summary-table}
mhsp_gpra_cols <- list(
  list("gpra_1a_target", "GPRA 1A Target"), list("gpra_1a_actual", "GPRA 1A Actual"),
  list("gpra_1b_target", "GPRA 1B Target"), list("gpra_1b_actual", "GPRA 1B Actual"),
  list("gpra_2a_target", "GPRA 2A Target"), list("gpra_2a_actual", "GPRA 2A Actual"),
  list("gpra_2b_target", "GPRA 2B Target"), list("gpra_2b_actual", "GPRA 2B Actual"),
  list("gpra_3a_target", "GPRA 3A Target"), list("gpra_3a_actual", "GPRA 3A Actual"),
  list("gpra_3b_target", "GPRA 3B Target"), list("gpra_3b_actual", "GPRA 3B Actual")
)

mhsp_summary <- bind_rows(lapply(mhsp_gpra_cols, function(x) {
  summarise_gpra_col(mhsp, x[[1]], x[[2]])
}))

kable(mhsp_summary, format = "html",
      caption = "Table 1. 2024 MHSP GPRA Targets and Actuals Summary")
```

---

## 4. SBMH

### 4.1 Import

```{r sbmh-import}
sbmh <- import_ipr_sheet("data/REVISED_SBMH_IPR_Synthetic.xlsx")
cat("SBMH rows:", nrow(sbmh), "| Columns:", ncol(sbmh), "\n")
head(sbmh)
```

### 4.2 Duplicate Grant IDs

```{r sbmh-dupes}
sbmh <- deduplicate_grants(sbmh, id_col = "Grant ID")
```

### 4.3 Cross-check Against APR Roster

```{r sbmh-crosscheck, message=FALSE}
sbmh_apr <- suppressMessages(
  read_excel("data/APRs_Revised_Tiers_Synthetic.xlsx", sheet = 2)
)
cat("APR reference rows (SBMH):", nrow(sbmh_apr), "\n")
crosscheck_grants(sbmh$`Grant ID`, sbmh_apr$grantee_number,
                  ipr_label = "SBMH IPR", apr_label = "SBMH APR")
```

---

## 5. SBMH GPRA Measures

```{r sbmh-clean-gpra, warning=FALSE}
sbmh <- sbmh |>
  clean_gpra_col("GPRA 1 (Hired) Target (Raw)",       "gpra_1_target") |>
  clean_gpra_col("GPRA 1 (Hired) Actual (Raw)",        "gpra_1_actual") |>
  clean_gpra_col("GPRA 2 (Retention) Target (Raw)",   "gpra_2_target") |>
  clean_gpra_col("GPRA 2 (Retention) Actual (Raw)",   "gpra_2_actual") |>
  clean_gpra_col("GPRA 4 (Attrition) Target (Raw)",   "gpra_4_target") |>
  clean_gpra_col("GPRA 4 (Attrition) Actual (Raw)",   "gpra_4_actual") |>
  clean_gpra_col("GPRA 5 (Students Served) Target (Raw)", "gpra_5_target") |>
  clean_gpra_col("GPRA 5 (Students Served) Actual (Raw)", "gpra_5_actual")
```

### 5.1 GPRA 1 -- Providers Hired

```{r sbmh-gpra1}
summarise_gpra_col(sbmh, "gpra_1_target", "GPRA 1 Target")
summarise_gpra_col(sbmh, "gpra_1_actual", "GPRA 1 Actual")
missing_grantees(sbmh, "gpra_1_actual")
```

### 5.2 GPRA 2 -- Provider Retention

```{r sbmh-gpra2}
summarise_gpra_col(sbmh, "gpra_2_target", "GPRA 2 Target")
summarise_gpra_col(sbmh, "gpra_2_actual", "GPRA 2 Actual")
missing_grantees(sbmh, "gpra_2_actual")
```

### 5.3 GPRA 3 -- Student-to-Provider Ratio

GPRA 3 is reported as a ratio string (e.g., "45000:12") rather than a raw count.
The `aggregate_ratios()` function parses and sums all numerators and denominators
to produce a program-wide ratio.

```{r sbmh-gpra3}
sbmh <- rename(sbmh,
  gpra_3_target = `GPRA 3 (Ratio) Target`,
  gpra_3_actual = `GPRA 3 (Ratio) Actual`
)

# Target
target_ratios <- aggregate_ratios(sbmh$gpra_3_target)
cat("GPRA 3 Target -- Total:", target_ratios$total_numerator, ":",
    target_ratios$total_denominator, "| Simplified:", target_ratios$simplified_ratio, "\n")

# Actual
actual_ratios <- aggregate_ratios(sbmh$gpra_3_actual)
cat("GPRA 3 Actual -- Total:", actual_ratios$total_numerator, ":",
    actual_ratios$total_denominator, "| Simplified:", actual_ratios$simplified_ratio, "\n")

# Grantees with missing ratio data
sbmh |> filter(is.na(gpra_3_actual)) |>
  select(any_of(c("Grant ID", "Grantee Name", "Region", "Current Status",
                   "Contact Person", "Contact Email")))
```

### 5.4 GPRA 4 -- Provider Attrition

```{r sbmh-gpra4}
# One grantee's GPRA 4 values are flagged as data entry errors; set to NA
# before cleaning (update Grant ID to match the real dataset)
# sbmh[sbmh$`Grant ID` == "S184H220226", c("gpra_4_target", "gpra_4_actual")] <- NA

summarise_gpra_col(sbmh, "gpra_4_target", "GPRA 4 Target")
summarise_gpra_col(sbmh, "gpra_4_actual", "GPRA 4 Actual")

# Flag outliers (attrition > 100 is likely a data entry error)
sbmh |>
  filter(gpra_4_target > 100 | gpra_4_actual > 100) |>
  select(any_of(c("Grant ID", "Grantee Name", "gpra_4_target", "gpra_4_actual")))
```

### 5.5 GPRA 5 -- Students Served

```{r sbmh-gpra5}
summarise_gpra_col(sbmh, "gpra_5_target", "GPRA 5 Target")
summarise_gpra_col(sbmh, "gpra_5_actual", "GPRA 5 Actual")
missing_grantees(sbmh, "gpra_5_actual")
```

### 5.6 SBMH Summary Table

```{r sbmh-summary-table}
sbmh_gpra_cols <- list(
  list("gpra_1_target", "GPRA 1 Target"), list("gpra_1_actual", "GPRA 1 Actual"),
  list("gpra_2_target", "GPRA 2 Target"), list("gpra_2_actual", "GPRA 2 Actual"),
  list("gpra_4_target", "GPRA 4 Target"), list("gpra_4_actual", "GPRA 4 Actual"),
  list("gpra_5_target", "GPRA 5 Target"), list("gpra_5_actual", "GPRA 5 Actual")
)

sbmh_summary <- bind_rows(lapply(sbmh_gpra_cols, function(x) {
  summarise_gpra_col(sbmh, x[[1]], x[[2]])
}))

kable(sbmh_summary, format = "html",
      caption = "Table 2. 2024 SBMH GPRA Targets and Actuals Summary")
```

---

## 6. Revision Follow-Up

```{r revision-summary}
bind_rows(
  revision_summary(mhsp, followup_col = "Follow-Up Needed", label = "MHSP"),
  revision_summary(sbmh, followup_col = "Follow- Up Needed (revisions)", label = "SBMH")
) |> kable(format = "html", caption = "Revision Follow-Up Summary")
```

```{r export-revisions}
# Export grantees who still need revisions
filtered_mhsp <- mhsp |>
  filter(grepl("Yes", `Revisions Required`, ignore.case = TRUE)) |>
  select(any_of(c("Grant ID", "Grantee Name", "Contact Person", "Contact Email",
                   "Region", "Current Status", "Follow-Up Needed",
                   "Revisions Required", "Revisions")))

filtered_sbmh <- sbmh |>
  filter(grepl("Yes", `Revisions Required?`, ignore.case = TRUE)) |>
  select(any_of(c("Grant ID", "Grantee Name", "Contact Person", "Contact Email",
                   "Region", "Current Status", "Follow- Up Needed (revisions)",
                   "Revisions Required?",
                   "Please provide more information about the revision")))

write_xlsx(
  list(MHSP = filtered_mhsp, SBMH = filtered_sbmh),
  path = "data/IPR_Grantees_Needing_Revisions.xlsx"
)
cat("Saved: data/IPR_Grantees_Needing_Revisions.xlsx\n")
```

---

## 7. Session Info

```{r session-info}
sessionInfo()
```
