---
title: "IPR GPRA Benchmark Tiers -- 2 GPRA Threshold"
author: "WCER School Mental Health Collaborative"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook:
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  html_document:
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes
---

## Overview

This notebook assigns GPRA benchmark tiers using the **2-GPRA threshold rule**:
a grantee is elevated to Tier 2 or Tier 3 only if **at least 2** GPRA measures
fall below the benchmark cutoff. This is the more conservative classification
compared to `02_IPR_GPRA_Benchmark_1Threshold.Rmd`.

**Tier definitions** (actual as a percentage of target):

| Tier | Criterion |
|------|-----------|
| Tier 1 | >= 75% of target (on track) |
| Tier 2 | 50-74% of target (needs attention) |
| Tier 3 | < 50% of target, or missing data (priority support) |

**Overall tier rule (2-threshold):** a grantee is elevated only when 2 or more
GPRA measures are off track; a single off-track measure defaults to Tier 1.

---

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
source("R/helpers.R")
load_packages(c("readxl", "tidyverse", "writexl", "knitr"))
```

---

## 2. Import and Clean

```{r import-clean, warning=FALSE}
mhsp <- import_ipr_sheet("data/REVISED_MHSP_IPR_Synthetic.xlsx") |>
  deduplicate_grants(id_col = "Grant ID") |>
  clean_gpra_col("GPRA 1A Target (Raw)", "gpra_1a_target") |>
  clean_gpra_col("GPRA 1A Actual (Raw)", "gpra_1a_actual") |>
  clean_gpra_col("GPRA 1B Target (Raw)", "gpra_1b_target") |>
  clean_gpra_col("GPRA 1B Actual (Raw)", "gpra_1b_actual") |>
  clean_gpra_col("GPRA 2A Target (Raw)", "gpra_2a_target") |>
  clean_gpra_col("GPRA 2A Actual (Raw)", "gpra_2a_actual") |>
  clean_gpra_col("GPRA 2B Target (Raw)", "gpra_2b_target") |>
  clean_gpra_col("GPRA 2B Actual (Raw)", "gpra_2b_actual") |>
  clean_gpra_col("GPRA 3A Target (Raw)", "gpra_3a_target") |>
  clean_gpra_col("GPRA 3A Actual (Raw)", "gpra_3a_actual") |>
  clean_gpra_col("GPRA 3B Target (Raw)", "gpra_3b_target") |>
  clean_gpra_col("GPRA 3B Actual (Raw)", "gpra_3b_actual") |>
  standardize_names()

sbmh <- import_ipr_sheet("data/REVISED_SBMH_IPR_Synthetic.xlsx") |>
  deduplicate_grants(id_col = "Grant ID") |>
  clean_gpra_col("GPRA 1 (Hired) Target (Raw)",        "gpra_1_target") |>
  clean_gpra_col("GPRA 1 (Hired) Actual (Raw)",         "gpra_1_actual") |>
  clean_gpra_col("GPRA 2 (Retention) Target (Raw)",    "gpra_2_target") |>
  clean_gpra_col("GPRA 2 (Retention) Actual (Raw)",    "gpra_2_actual") |>
  clean_gpra_col("GPRA 4 (Attrition) Target (Raw)",    "gpra_4_target") |>
  clean_gpra_col("GPRA 4 (Attrition) Actual (Raw)",    "gpra_4_actual") |>
  clean_gpra_col("GPRA 5 (Students Served) Target (Raw)", "gpra_5_target") |>
  clean_gpra_col("GPRA 5 (Students Served) Actual (Raw)", "gpra_5_actual") |>
  rename(gpra_3_target = `GPRA 3 (Ratio) Target`,
         gpra_3_actual = `GPRA 3 (Ratio) Actual`) |>
  standardize_names() |>
  mutate(
    gpra3_target_decimal = ratio_to_decimal(gpra3_target),
    gpra3_actual_decimal = ratio_to_decimal(gpra3_actual)
  )

# Recode NA sentinel strings in all gpra columns
recode_gpra_na <- function(df) {
  df |> mutate(across(
    contains("gpra", ignore.case = TRUE),
    ~ na_if(gsub("(N/A|n/a|n/a, 999|na|NA)", "", as.character(.)), "")
  ))
}
mhsp <- recode_gpra_na(mhsp)
sbmh <- recode_gpra_na(sbmh)

cat("MHSP:", nrow(mhsp), "grantees | SBMH:", nrow(sbmh), "grantees\n")
```

---

## 3. Revision Tiers

```{r revision-tiers}
mhsp <- mhsp |>
  mutate(revision_tier = case_when(
    followup_needed == TRUE & revised == TRUE & is.na(conf_email_needed) ~ "Tier 2",
    conf_email_needed == TRUE                                            ~ "Tier 1",
    followup_needed == TRUE & is.na(revised) & is.na(conf_email_needed) ~ "Tier 3",
    TRUE ~ NA_character_
  ))

sbmh <- sbmh |>
  mutate(revision_tier = case_when(
    followup_needed == TRUE & revised == TRUE & is.na(conf_email_needed) ~ "Tier 2",
    conf_email_needed == TRUE                                            ~ "Tier 1",
    followup_needed == TRUE & is.na(revised) & is.na(conf_email_needed) ~ "Tier 3",
    TRUE ~ NA_character_
  ))

bind_rows(
  mhsp |> count(revision_tier) |> mutate(pct = round(n / sum(n) * 100, 1), Program = "MHSP"),
  sbmh |> count(revision_tier) |> mutate(pct = round(n / sum(n) * 100, 1), Program = "SBMH")
) |> kable(format = "html", caption = "Revision Tier Distribution")
```

---

## 4. GPRA Benchmark Tiers -- 2-Threshold Rule

```{r benchmark-tiers-sbmh}
sbmh <- sbmh |>
  mutate(across(c(gpra1_actual, gpra1_target, gpra2_actual, gpra2_target,
                  gpra3_actual_decimal, gpra3_target_decimal,
                  gpra4_actual, gpra4_target, gpra5_actual, gpra5_target),
                as.numeric),
         gpra1_tier = mapply(benchmark_tier, gpra1_actual, gpra1_target),
         gpra2_tier = mapply(benchmark_tier, gpra2_actual, gpra2_target),
         gpra3_tier = mapply(benchmark_tier, gpra3_actual_decimal, gpra3_target_decimal),
         gpra4_tier = mapply(benchmark_tier, gpra4_actual, gpra4_target),
         gpra5_tier = mapply(benchmark_tier, gpra5_actual, gpra5_target)) |>
  mutate(
    overall_tier_num = pmap_dbl(
      list(gpra1_tier, gpra2_tier, gpra3_tier, gpra4_tier, gpra5_tier),
      \(...) assign_overall_tier(..., tiers_required = 2)   # 2-threshold
    ),
    benchmark_tier = case_when(
      overall_tier_num == 3 ~ "Tier 3",
      overall_tier_num == 2 ~ "Tier 2",
      overall_tier_num == 1 ~ "Tier 1",
      TRUE ~ NA_character_
    )
  )
```

```{r benchmark-tiers-mhsp}
mhsp <- mhsp |>
  mutate(across(c(gpra1_a_actual, gpra1_a_target, gpra1_b_actual, gpra1_b_target,
                  gpra2_a_actual, gpra2_a_target, gpra2_b_actual, gpra2_b_target,
                  gpra3_a_actual, gpra3_a_target, gpra3_b_actual, gpra3_b_target),
                as.numeric),
         gpra1_a_tier = mapply(benchmark_tier, gpra1_a_actual, gpra1_a_target),
         gpra1_b_tier = mapply(benchmark_tier, gpra1_b_actual, gpra1_b_target),
         gpra2_a_tier = mapply(benchmark_tier, gpra2_a_actual, gpra2_a_target),
         gpra2_b_tier = mapply(benchmark_tier, gpra2_b_actual, gpra2_b_target),
         gpra3_a_tier = mapply(benchmark_tier, gpra3_a_actual, gpra3_a_target),
         gpra3_b_tier = mapply(benchmark_tier, gpra3_b_actual, gpra3_b_target)) |>
  mutate(
    overall_tier_num = pmap_dbl(
      list(gpra1_a_tier, gpra1_b_tier, gpra2_a_tier, gpra2_b_tier,
           gpra3_a_tier, gpra3_b_tier),
      \(...) assign_overall_tier(..., tiers_required = 2)   # 2-threshold
    ),
    benchmark_tier = case_when(
      overall_tier_num == 3 ~ "Tier 3",
      overall_tier_num == 2 ~ "Tier 2",
      overall_tier_num == 1 ~ "Tier 1",
      TRUE ~ NA_character_
    )
  )
```

---

## 5. Tier Distribution Results

```{r tier-results}
cat("SBMH Benchmark Tier Distribution (2-Threshold):\n")
sbmh |> count(benchmark_tier) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  kable(format = "html")

cat("\nMHSP Benchmark Tier Distribution (2-Threshold):\n")
mhsp |> count(benchmark_tier) |>
  mutate(pct = round(n / sum(n) * 100, 1)) |>
  kable(format = "html")
```

### Per-GPRA Tier Distributions

```{r gpra-tier-breakdown}
tier_cols_sbmh <- c("gpra1_tier","gpra2_tier","gpra3_tier","gpra4_tier","gpra5_tier")
lapply(tier_cols_sbmh, \(col) {
  sbmh |> count(!!sym(col)) |> mutate(pct = round(n / sum(n) * 100, 1), gpra = col)
}) |> bind_rows() |> kable(format = "html", caption = "SBMH Per-GPRA Tier Breakdown")

tier_cols_mhsp <- c("gpra1_a_tier","gpra1_b_tier","gpra2_a_tier","gpra2_b_tier",
                    "gpra3_a_tier","gpra3_b_tier")
lapply(tier_cols_mhsp, \(col) {
  mhsp |> count(!!sym(col)) |> mutate(pct = round(n / sum(n) * 100, 1), gpra = col)
}) |> bind_rows() |> kable(format = "html", caption = "MHSP Per-GPRA Tier Breakdown")
```

---

## 6. Export

```{r export}
write_xlsx(
  list(SBMH = sbmh, MHSP = mhsp),
  path = "data/IPR_BenchmarkTiers_2Threshold.xlsx"
)
cat("Saved: data/IPR_BenchmarkTiers_2Threshold.xlsx\n")
```

---

## 7. Session Info

```{r session-info}
sessionInfo()
```
